### 2021-07-22 paper/paper.tex
first draft of assessment

\todo{TODO %TODO
\begin{itemize}
  \item Why we created \trixi
  \item Overview of the paper \etc
\end{itemize}}

@@

\todo{TODO: Andrew, please describe the setups that you used} %TODO

@@

\cite{byrne2021mpi}

\todo{TODO: p4est \cite{burstedde2011p4est} (HDF5)} %TODO

@@

\todo{TODO} %TODO

@@

\todo{TODO} %TODO

@@

Multi-physics with 350 lines of code \cite{schlottkelakemper2021purely}
\todo{TODO} %TODO

\subsubsection{\trixi can be extended from outside}

\todo{TODO} %TODO

@@

\todo{TODO %TODO
\begin{itemize}
  \item Compilation time can be annoying (serial compiler!) but also allows cool features and performance
  \item MPI: To be evaluated but the basic tools are already there \cite{byrne2021mpi}
\end{itemize}}

@@

\todo{TODO} %TODO

@@

\todo{TODO: Add your funding sources (if any)} %TODO

>>>>

\todo{TODO %TODO Introduction
\begin{itemize}
  \item Why we created \trixi
  \item Overview of the paper \etc
\end{itemize}}

@@

\todo{TODO: Andrew, please describe the setups that you used} %TODO Andrew

@@

We have written a Julia wrapper\footnote{\url{https://github.com/trixi-framework/P4est.jl}}
of the C library \texttt{p4est} \cite{burstedde2011p4est}. In our experience,
it is relatively easy to do so in Julia and the package manager in combination
with BinaryBuilder.jl makes it convenient to distribute the necessary binaries.
In particular, users do not have to compile C libraries on their own system
and the binaries are available on all major platforms including Linux, Mac OS,
and Windows.

Similarly, there is no need to compile HDF5 on a user system due to the
availability of HDF5.jl. There is still work to be done how to improve the
experience in combination with MPI on high performance computing (HPC) clusters,
but some promising approaches exist \cite{byrne2021mpi}.

@@

Due to Julia's package manager, the cost of using external dependencies is low
enough to not be a problem. Traditionally, many scientific simulation codes tend
to reduce the number of external dependencies as much as possible due to the
complexity of handling different versions and making everything work together.
Such an approach often leads to a lot of code duplication. For example, many
CFD codes implement their own time integration methods, arguing that the spatial
part is much more complex. While this is often true, it is still more efficient
in terms of human workload to be able to reuse existing implementations. This
allows expert on time integration methods to develop specialized algorithms and
implement them in open source software while practitioners or researchers focusing
on spatial semidiscretizations ca benefit immediately by changing only one line
of code. For example, optimized time integration methods were developed in
\cite{ranocha2021optimized} and implemented in OrdinaryDiffEq.jl. Thus, they
can be used with \trixi by changing a single line of code. In contrast, researchers
working without external dependencies would need to use a day or two to digest
all the details and implement the methods in their own codes, making it much
less likely that they will do this and benefit from recent algorithmic developments.

An additional example is given by automatic differentiation. Due to dispatch,
it is conveniently possible to use forward mode automatic differentiation
\cite{revels2016forward} to compute Jacobians of semidiscretizations of nonlinear
conservation laws with \trixi or to differentiate through a complete simulation
including time integration methods from OrdinaryDiffEq.jl.

@@

Due to the package manager, \trixi and all related ODE and visualization tools
can be installed with a single command, reducing the overhead of exploring the
code. In addition, Julia's expressiveness and the ability to work with a restricted
subset of all possible features have enabled more than 18 students to use \trixi
for their coursework or theses. In our experience, the time required to get
started with traditional monolithic code bases is so long that students either
choose to build their own implementations specialized on their tasks or can only
work on toy problems since the remaining time is rather restricted.

@@

Due to Julia's high-level programming approach, dynamic typing, and multiple
dispatch, it is easy to combine existing functionality efficiently. For example,
the single-physics solvers for hyperbolic PDEs of \trixi were extended to
a multi-physics setup for the compressible Euler equations with self-gravity
with roughly 350 lines of code \cite{schlottkelakemper2021purely}.
In addition, \trixi can be extended from the outside without modifying the main
source code, which makes it easy to set up new simulation approaches and analyze
existing ones, \eg by fluctuation simulations \cite{ranocha2021preventing}.

@@

\subsubsection{Compilation times can be annoying}

Since Julia uses a serial just-ahead-of-time compiler that does currently not cache
code between different Julia sessions, the initial compilation time can be
annoying. For example, the time to finish the first simulation in a fresh Julia
session and plot the results on a notebook can take between 30 seconds and a
minute. Having compiled the code, the second simulation and plot take less than
0.05 seconds. Thus, switching to Julia requires adapting the workflow compared
to other languages such as keeping a REPL session active for a longer time and
using packages such as Revise.jl\footnote{\url{https://github.com/timholy/Revise.jl}}.
This is particularly problematic in HPC environments. There are tools such as
PackageCompiler.jl\footnote{\url{https://github.com/JuliaLang/PackageCompiler.jl}}
that can help with these tasks but a good workflow usable for development and
deployment still has to be found.

\subsubsection{MPI-based parallel simulations need to be established}

MPI: To be evaluated but the basic tools are already there \cite{byrne2021mpi}

\todo{TODO: Michael, could you please write something here? You are much more
experienced with MPI and HPC} %TODO Michael

@@

\todo{TODO %TODO Summary
\begin{itemize}
  \item Learning a new programming language has some overhead but it pays off
\end{itemize}}

@@

\todo{TODO: Add your funding sources that need to be here (if any)} %TODO Funding
